Numerical simulations of the Cordilleran ice sheet through the last glacial cycle

### 2015-06-20 cordillera-cycle.tex
Add corrections from Arjen and Martin (part 1).

  Despite more than a century of geological observations, the Cordilleran ice
  sheet of North America remains poorly understood in terms of its former
  extent, volume and dynamics. Although geomorphological evidence is abundant,
  its complexity is such that whole ice-sheet reconstructions of advance and
  retreat patterns are lacking. Here we use a numerical ice sheet model
  calibrated against field-based evidence to attempt a quantitative
  reconstruction of the Cordilleran ice sheet history through the last glacial
  cycle. A series of simulations is driven by time-dependent temperature
  offsets from six proxy records located around the globe. Although this
  approach reveals large variations in model response to evolving climate
  forcing, all simulations produce two major glaciations during
  marine oxygen isotope stages~4 (61.9--56.5\,ka) and~2
  (23.2--16.8\,ka). The timing of glaciation is
  better reproduced using temperature reconstructions from Greenland and
  Antarctic ice cores than from regional oceanic sediment cores. During most of
  the last glacial cycle, the modelled ice cover is discontinuous and
  restricted to high mountain areas. However, widespread precipitation over the
  Skeena Mountains favours the persistence of a central ice dome throughout the
  glacial cycle. It acts as a nucleation centre before the last glacial maximum
  and hosts the last remain of Cordilleran ice until the middle Holocene
  (6.6--6.2\,ka).

@@

Our understanding of the Cordilleran glaciation history prior to the LGM is
even more fragmentary
    \citep{Barendregt.Irving.1998, Kleman.etal.2010, Rutter.etal.2012},
although it is clear that maximum glaciation of the Cordilleran ice sheet
predates the last glacial cycle
    \citep{Hidy.etal.2013}.
In parts of the Yukon Territory and Alaska, the distribution of tills
    \citep{Turner.etal.2013}
and dated glacial erratics
    \citep{Ward.etal.2007, Ward.etal.2008, Briner.Kaufman.2008,
           Stroeven.etal.2010, Stroeven.etal.2014}
indicate an extensive Marine Oxygen Isotope Stage (MIS)~4 glaciation, the
evidence of which have also been found in the Puget Lowland \citep{Troost.2014}.
Landforms in the interior regions include flow sets that are likely
older than the LGM
    \citep[Fig.~2]{Kleman.etal.2010},
but their absolute age remains uncertain.

\begin{figure}
  \includegraphics{locmap}
  \caption{Relief map of northern North America showing a reconstruction of the
           areas once covered by the Cordilleran (CIS), Laurentide (LIS),
           Innuitian (IIS) and Greenland (GIS) ice sheets during the last
           18\,\unit{\chem{^{14}C}\,ka} \citep[21.4\,cal\,ka,][]{Dyke.2004}.
           The rectangular box denotes the location of the
           modelling domain used in this study. Major mountain ranges covered
           by the ice sheet include the Alaska Range (AR), the Wrangell and
           St.-Elias mountains (WSEM), the Selwyn and Mackenzie mountains
           (SMKM), the Skeena Mountains (SM), the Coast and Rocky
           Mountains (CRM) and the North Cascades (NC). Major depressions
           include the Liard Lowland (LL), the Interior Plateau of British
           Columbia (IP) and the Puget Lowland (PL). The background
           map consists of ETOPO1 \citep{Amante.Eakins.2009} and Natural Earth
           Data \citep{Patterson.Kelso.2015}.}
  \label{fig:locmap}
\end{figure}

In contrast, evidence for the deglaciation history of the Cordilleran
ice sheet since the LGM is considerable, albeit mostly at a regional scale.
Geomorphological evidence from south-central British Columbia indicates a rapid
deglaciation, including an early emergence of the elevated areas while thin,
stagnant ice still covered the surrounding lowlands
    \citep{Fulton.1967, Fulton.1991, Margold.etal.2011, Margold.etal.2013a}.
This model, although credible, may not apply in all areas of the Cordilleran
ice sheet
    \citep{Margold.etal.2013}.
Although solid evidence for late-glacial glacier re-advances have been found in
the Coast, Columbia and Rocky mountains
    \citep{Clague.etal.1997, Friele.Clague.2002, Friele.Clague.2002a,
           Kovanen.2002, Kovanen.Easterbrook.2002, Lakeman.etal.2008,
           Menounos.etal.2008},
it appears to be sparser than for formerly glaciated regions surrounding
the North Atlantic
    \citep[e.g.,][]{Sissons.1979, Lundqvist.1987,
                    Ivy-Ochs.etal.1999, Stea.etal.2011}.
Nevertheless, recent oxygen isotope measurements from Gulf of Alaska sediments
reveals a climatic evolution highly correlated to that of Greenland during this
period, including a distinct late glacial cold reversal between 14.1 and
11.7\,ka \citep{Praetorius.Mix.2014}.

In general, the topographic complexity of the North American Cordillera and its
effect on glacial history have inhibited the reconstruction of ice sheet-wide
glacial advance and retreat patterns such as
those available for the Fennoscandian and Laurentide ice sheets
     \citep{Boulton.etal.2001, Dyke.Prest.1987, Dyke.etal.2003,
            Kleman.etal.1997, Kleman.etal.2010, Stroeven.etal.inreview}.
Here, we use a numerical ice sheet model \citep{PISM-authors.2015},
calibrated against field-based evidence, to perform a quantitative
reconstruction of the Cordilleran ice sheet history through the last glacial
cycle, and
analyse some of the long-standing questions related to its evolution:

\begin{itemize}
  \item How much ice was locked in the Cordilleran ice sheet during the LGM?
  \item What was the scale of glaciation prior to the LGM?
  \item Which were the primary dispersal centres? Do they reflect stable or
    ephemeral configurations?
  \item How rapid was the last deglaciation? Did it include late glacial
    standstills or readvances?
\end{itemize}

Although numerical ice sheet modelling has been established as a useful tool to
improve our understanding of the Cordilleran ice sheet
    (\citealp[p.~227]{Jackson.Clague.1991}; \citealp{Robert.1991},
     \citealp{Marshall.etal.2000}),
the ubiquitously mountainous
topography of the region has presented two major challenges to its application.
First, only recent development of numerical ice sheet models and underlying
scientific computing tools \citep{Bueler.Brown.2009, Balay.etal.2015} has
allowed for high-resolution numerical modelling of glaciers and ice sheets on
mountainous terrain
over millennial time scales \citep[e.g.,][]{Golledge.etal.2012}. Second, the
complex topography of the North American Cordillera also induces strong
geographic variations in temperature and precipitation, thus requiring the use
of high-resolution climate forcing fields as an input to an ice sheet model
\citep{Seguinot.etal.2014}. However, evolving climate conditions over the last
glacial cycle are subject to considerable uncertainty and still lie beyond the
computational reach of atmosphere circulation models.

@@

Ice rheology depends on temperature and water content through an enthalpy
formulation \citep{Aschwanden.etal.2012}. Surface air temperature derived from
the climate forcing (Sect.~\ref{sec:atm}) provides the upper boundary
condition to the ice enthalpy model. Temperature is computed subglacially to
a~depth of 3\,km,
where it is conditioned by a lower boundary geothermal heat flux of
70\,\unit{mW\,m^{-2}}. Although this uniform value does not account for the
high spatial geothermal variability in the region
\citep{Blackwell.Richards.2004}, it is, on average, representative of available
heat flow measurements. In the low-resolution simulations, the vertical grid
consists of up to 51~enthalpy layers in the ice sheet and 31~temperature layers
in the bedrock. In the high-resolution simulations, up to 101~ice layers and
61~bedrock layers are used.

A~pseudo-plastic sliding law,
\begin{equation}
    \label{eqn:pseudoplastic}
    \vec{\tau}_b = -\tau_c \frac{\vec{v}_b}{{v_{th}}^q\,|\vec{v}_b|^{1-q}} \,,
\end{equation}
relates the bed-parallel shear stresses, $\vec{\tau}_b$, to the sliding
velocity, $\vec{v}_b$ (Table~\ref{tab:params}).
The yield stress, $\tau_c$,
is modelled using the Mohr--Coulomb criterion,
\begin{equation}
   \tau_c = c_0 + N\,\tan{\phi} \,,
\end{equation}
where cohesion, $c_0$, is assumed to be zero. The friction angle, $\phi$,
varies from 15 to 45\degree as a piecewise-linear function of modern bed
elevation, with the lowest value occuring below modern sea level (0\,m above
sea level, m~a.s.l.) and the highest value occuring above the generalised
elevation of the highest shorelines
\citep[200\,m~a.s.l.,][Fig.~5]{Clague.1981}, thus accounting for
a~weakening of till associated with the presence of marine sediments
(cf. \citealp{Martin.etal.2011}; \citealp[supplement]{Aschwanden.etal.2013};
\citealp{PISM-authors.2015}). Effective pressure, $N$,
is related to the ice overburden stress, $\rho gh$, and the modelled amount of
subglacial water, using a formula derived from laboratory experiments with till
extracted from the base of Ice Stream B in West Antarctica
\citep{Tulaczyk.etal.2000, Bueler.Pelt.2014},
\begin{equation}
    N = \delta \rho gh \, 10^{(e_0/C_c) (1 - (W/W_{max}))} \,,
\end{equation}
where $\delta$ is chosen as 0.02, $e_0$ is a measured reference void ratio and
$C_c$ is a measured compressibility coefficient (Table~\ref{tab:params}). The
amount of water at the base, $W$, varies from zero to $W_{max}=2$\,m, a
threshold above which any additional melt water is assumed to drain off
instantaneously. Finally, the bedrock topography responds to ice load
following a bedrock deformation model that includes local isostasy,
elastic lithosphere flexure and viscous astenosphere deformation in an infinite
half-space
\citep[Table~\ref{tab:params};][]{Lingle.Clark.1985,Bueler.etal.2007}.
A relatively low viscosity value of $\nu_m = 1\times10^{19}$\,\unit{Pa\,s} is
used for the astenosphere in accordance to the results from regional glacial
isostatic adjustment modelling at the northern Cascadia subduction zone
\citep{James.etal.2009}.

Ice-shelf calving is computed by a double criterion. First, a
physically-realistic calving flux is computed based on eigenvalues of the
horizontal strain rate tensor \citep{Winkelmann.etal.2011,
Levermann.etal.2012}. This allows outlet glaciers to advance in confined
embayments, but prevents extensive ice-shelves from building up in the open ocean.
Second, floating ice thinner than 50\,m is systematically calved off. A subgrid
scheme \citep{Albrecht.etal.2011} allows for a continuous migration of the
calving front. Such treatment of calving has been applied to the Antarctic ice
sheet and has shown to produce a realistic calving front positon for many of
the present-day ice shelves \citep{Martin.etal.2011}.

@@

\begin{figure}
  \includegraphics{atm}
  \caption{Monthly mean near-surface air temperature, precipitation, and
           standard deviation of daily mean temperature for January and July
           from the North American Regional Reanalysis
           \citep[NARR;][]{Mesinger.etal.2006}, used to force the surface mass
           balance (PDD) component of the ice sheet model. Note the
           strong contrasts in seasonality, timing of the precipitation peak,
           and temperature variability over the model domain, notably between
           coastal and inland regions.}
  \label{fig:atm}
\end{figure}

@@

The PDD computation accounts for stochastic temperature variations by assuming
a normal temperature distribution of standard deviation $\sigma$ aroung the
expected value $T_m$. It is expressed by an error-function formulation
\citep{Calov.Greve.2005},
\begin{equation}
    \label{eqn:calovgreve}
    \mathrm{PDD} = \int_{t_1}^{t_2} \mathrm{d}t
        \left[\frac{\sigma}{\sqrt{2\pi}}
                \exp\left({-\frac{T_{m}^2}{2\sigma^2}}\right)
              + \frac{T_{m}}{2} \, \mathrm{erfc}
                \left(-\frac{T_{m}}{\sqrt{2}\sigma}\right)\right] \,,
\end{equation}
which is numerically approximated using week-long sub-intervals. In order to
account for the effects of spatial and seasonal variations of temperature
variability \citep{Seguinot.2013}, $\sigma$ is computed from NARR daily
temperature values from 1979 to 2000 \citep{Mesinger.etal.2006},
including variability associated with the seasonal cycle
itself \citep[Fig.~\ref{fig:atm}; i.e. not as in][]{Seguinot.Rogozhina.2014}.
Degree-day factors for snow and ice melt are derived from
mass-balance measurements on contemporary glaciers from the Coast Mountains and
Rocky Mountains in British Columbia
\citep[Table~\ref{tab:params};][]{Shea.etal.2009}.

@@

\begin{table*}[t]
  \caption{Palaeo-temperature proxy records and scaling parameters yielding
           temperature offset time-series used to force the ice sheet model
           through the last glacial cycle (Fig.~\ref{fig:timeseries}). $f$
           corresponds to the scaling factor adopted to yield last glacial
           maximum ice limits in the vicinity of mapped end moraines, and
           $T_{[32, 22]}$ refers to the resulting mean temperature anomaly
           during the period -32 to~-22~ka after scaling.}
  \label{tab:records}
  {\begin{tabular}{l|ccc|ccc|l}
    \tophline
    Record & Latitude & Longitude & Elev. & Proxy & $f$ & $T_{[32, 22]}$
           & Reference\\
    & & & (m~a.s.l.) & & & (K) & \\
    \middlehline
    GRIP     &  72{\degree} 35' N  % 72.58 (decimal)
             &  37{\degree} 38' W  % 37.64 (decimal)
             & 3238\,m
             & \chem{\delta^{18}O}
             & 0.37 & -5.8{\degree}C  % -16.4126 (before scaling)
             & \citet{Dansgaard.etal.1993} \\

    NGRIP    &  75{\degree} 06' N  % 75.10
             &  42{\degree} 19' W  % 42.32
             & 2917\,m
             & \chem{\delta^{18}O}
             & 0.24 & -6.5{\degree}C  % -26.7098
             & \citet{Andersen.etal.2004} \\

    EPICA    &  75{\degree} 06' S  % 75.1
             & 123{\degree} 21' E  % 123.35
             & 3233\,m
             & \chem{\delta^{18}O}
             & 0.64 & -5.9{\degree}C  % -9.2055
             & \citet{Jouzel.etal.2007} \\

    Vostok   &  78{\degree} 28' S  % 78.8
             & 106{\degree} 50' E  % 106.8
             & 3488\,m
             & \chem{\delta^{18}O}
             & 0.74 & -5.9{\degree}C  % -7.9550
             & \citet{Petit.etal.1999} \\

    ODP~1012 &  32{\degree} 17' N
             & 118{\degree} 23' W
             & -1772\,m
             & \chem{U^{K'}_{37}}
             & 1.61 & -6.1{\degree}C  % -3.7889
             & \citet{Herbert.etal.2001} \\

    ODP~1020 &  41{\degree} 00' N
             & 126{\degree} 26' W
             & -3038\,m
             & \chem{U^{K'}_{37}}
             & 1.20 & -6.0{\degree}C  % -5.0000
             & \citet{Herbert.etal.2001} \\
    \bottomhline
  \end{tabular}}
  \belowtable{}
\end{table*}

@@

While simulations driven by the GRIP and the two Antarctic palaeo-temperature
records attain a last ice volume maximum between 19.1 and 16.8\,ka, those
informed by the NGRIP and the two oceanic palaeo-temperature records attain their
maximum ice volumes thousands of years earlier. Moreover, the ODP~1012 run
yields a rapid deglaciation of the modelled area prior to 17\,ka. The ODP~1020
simulation predicts an early maximum in ice volume at 23.2\,ka, followed by
slower deglaciation than modelled using the other palaeo-temperature records.
Finally, model runs forced by the Antarctic palaeo-temperature records result in a
rapid and uninterrupted deglaciation after the LGM, whereas the
simulation driven by the GRIP palaeo-temperature record also results in a rapid
deglaciation but in three steps, separated by two periods of ice sheet regrowth
(Fig.~\ref{fig:timeseries}).

@@

During the MIS~3 ice volume minima, a central ice cap persists over the Skeena
Mountains (Fig.~\ref{fig:snapshots}, middle panels). Although this ice cap is
present in all simulations, its dimensions depend sensitively on the choice of
the applied palaeo-temperature record. Modelled ice volume minima spread over a
wide range between 1.69 and 2.88\,m~s.l.e. (Table~\ref{tab:extrema}).

Modelled ice sheet geometries during the LGM (MIS~2; Fig.~\ref{fig:snapshots},
lower panels) invariably include a ca. 1500\,km-long central divide located
about 3500\,m~a.s.l. along the spine of the Rocky Mountains. Although the
similatity of modelled ice extents is a direct result of the choice of scaling
factors applied to different palaeo-temperature proxy records
(Table~\ref{tab:records}), it is interesting to note that modelled maximum ice
volumes also fall within a tight range of 8.40 to 8.91\,m~s.l.e.
(Table~\ref{tab:extrema}).

@@

Independently of the palaeo-temperature records
used to force the ice sheet model, our simulations consistently produce two
glacial maxima during the last glacial cycle. The first maximum configuration
is obtained during MIS~4 (61.9--56.5\,ka) and the second during MIS~2
(23.2--16.8\,ka; Figs.~\ref{fig:timeseries}, \ref{fig:snapshots};
Table~\ref{tab:extrema}). These events broadly correspond in timing to the
Gladstone
(MIS~4) and McConnell (MIS~2) glaciations documented by geological evidence for
the northern sector of the Cordilleran ice sheet
    \citep{Duk-Rodkin.etal.1996, Ward.etal.2007,
           Stroeven.etal.2010, Stroeven.etal.2014},
and to the Fraser Glaciation (MIS~2) documented for its southern sector
    \citep{Porter.Swanson.1998, Margold.etal.2014}.
There is stratigraphical evidence for glaciations older than the Fraser
Glaciation in British Columbia \citep{Clague.Ward.2011} and in the Puget
Lowland \citep{Troost.2014}, but their extent and
timing are still highly conjectural
    \citep[perhaps MIS~4 or early MIS~3; e.g.,][]{Cosma.etal.2008}.

@@

During MIS~2, the modelled total ice volume peaks at 8.01\,m~s.l.e. in the GRIP
simulation
and at 8.77\,m~s.l.e. in the EPICA simulation.

>>>>

  Despite more than a century of geological observations, the Cordilleran ice
  sheet of North America remains poorly understood in terms of its former
  extent, volume and dynamics. Although geomorphological evidence is abundant,
  its complexity is such that whole ice-sheet reconstructions of advance and
  retreat patterns are lacking. Here we use a numerical ice sheet model
  calibrated against field-based evidence to attempt a quantitative
  reconstruction of the Cordilleran ice sheet history through the last glacial
  cycle. A series of simulations is driven by time-dependent temperature
  offsets from six proxy records located around the globe. Although this
  approach reveals large variations in model response to evolving climate
  forcing, all simulations produce two major glaciations during
  marine oxygen isotope stages~4 (61.9--56.5\,ka) and~2
  (23.2--16.8\,ka). The timing of glaciation is
  better reproduced using temperature reconstructions from Greenland and
  Antarctic ice cores than from regional oceanic sediment cores. During most of
  the last glacial cycle, the modelled ice cover is discontinuous and
  restricted to high mountain areas. However, widespread precipitation over the
  Skeena Mountains favours the persistence of a central ice dome throughout the
  glacial cycle. It acts as a nucleation centre before the Last Glacial Maximum
  and hosts the last remains of Cordilleran ice until the middle Holocene
  (6.6--6.2\,ka).

@@

Our understanding of the Cordilleran glaciation history prior to the LGM is
even more fragmentary
    \citep{Barendregt.Irving.1998, Kleman.etal.2010, Rutter.etal.2012},
although it is clear that the Pleistocene maximum extent of the Cordilleran
ice sheet predates the last glacial cycle
    \citep{Hidy.etal.2013}.
In parts of the Yukon Territory and Alaska, and in the Puget Lowland,
the distribution of tills
    \citep{Turner.etal.2013, Troost.2014}
and dated glacial erratics
    \citep{Ward.etal.2007, Ward.etal.2008, Briner.Kaufman.2008,
           Stroeven.etal.2010, Stroeven.etal.2014}
indicate an extensive Marine Oxygen Isotope Stage (MIS)~4 glaciation.
Landforms in the interior regions include flow sets that are likely
older than the LGM
    \citep[Fig.~2]{Kleman.etal.2010},
but their absolute age remains uncertain.

\begin{figure}
  \includegraphics{locmap}
  \caption{Relief map of northern North America showing a reconstruction of the
           areas once covered by the Cordilleran (CIS), Laurentide (LIS),
           Innuitian (IIS), and Greenland (GIS) ice sheets during the last
           18\,\unit{\chem{^{14}C}\,ka} \citep[21.4\,cal\,ka,][]{Dyke.2004}.
           The rectangular box denotes the location of the
           modelling domain used in this study. Major mountain ranges covered
           by the ice sheet include the Alaska Range (AR), the Wrangell and
           St.-Elias mountains (WSEM), the Selwyn and Mackenzie mountains
           (SMKM), the Skeena Mountains (SM), the Coast Mountains (CM), the
           Columbia and Rocky
           Mountains (CRM), and the North Cascades (NC). Major depressions
           include the Liard Lowland (LL), the Interior Plateau of British
           Columbia (IP), and the Puget Lowland (PL). The background
           map consists of ETOPO1 \citep{Amante.Eakins.2009} and Natural Earth
           Data \citep{Patterson.Kelso.2015}.}
  \label{fig:locmap}
\end{figure}

In contrast, evidence for the deglaciation history of the Cordilleran
ice sheet since the LGM is considerable, albeit mostly at a regional scale.
Geomorphological evidence from south-central British Columbia indicates a rapid
deglaciation, including an early emergence of elevated areas while thin,
stagnant ice still covered the surrounding lowlands
    \citep{Fulton.1967, Fulton.1991, Margold.etal.2011, Margold.etal.2013a}.
This model, although credible, may not apply in all areas of the Cordilleran
ice sheet
    \citep{Margold.etal.2013}.
Although solid evidence for late-glacial glacier re-advances have been found in
the Coast, Columbia and Rocky mountains
    \citep{Clague.etal.1997, Friele.Clague.2002, Friele.Clague.2002a,
           Kovanen.2002, Kovanen.Easterbrook.2002, Lakeman.etal.2008,
           Menounos.etal.2008},
it appears to be sparser than for formerly glaciated regions surrounding
the North Atlantic
    \citep[e.g.,][]{Sissons.1979, Lundqvist.1987,
                    Ivy-Ochs.etal.1999, Stea.etal.2011}.
Nevertheless, recent oxygen isotope measurements from Gulf of Alaska sediments
reveal a climatic evolution highly correlated to that of Greenland during this
period, including a distinct Late Glacial cold reversal between 14.1 and
11.7\,ka \citep{Praetorius.Mix.2014}.

In general, the topographic complexity of the North American Cordillera and its
effect on glacial history have inhibited the reconstruction of ice sheet-wide
glacial advance and retreat patterns such as
those available for the Fennoscandian and Laurentide ice sheets
     \citep{Boulton.etal.2001, Dyke.Prest.1987, Dyke.etal.2003,
            Kleman.etal.1997, Kleman.etal.2010, Stroeven.etal.inreview}.
Here, we use a numerical ice sheet model \citep{PISM-authors.2015},
calibrated against field-based evidence, to perform a quantitative
reconstruction of the Cordilleran ice sheet evolution through the last glacial
cycle, and
analyse some of the long-standing questions related to its evolution:

\begin{itemize}
  \item How much ice was locked in the Cordilleran ice sheet during the LGM?
  \item What was the scale of glaciation prior to the LGM?
  \item Which were the primary dispersal centres? Do they reflect stable or
    ephemeral configurations?
  \item How rapid was the last deglaciation? Did it include Late Glacial
    standstills or readvances?
\end{itemize}

Although numerical ice sheet modelling has been established as a useful tool to
improve our understanding of the Cordilleran ice sheet
    (\citealp[p.~227]{Jackson.Clague.1991}; \citealp{Robert.1991};
     \citealp{Marshall.etal.2000}),
the ubiquitously mountainous
topography of the region has presented two major challenges to its application.
First, only recent developments in numerical ice sheet models and underlying
scientific computing tools \citep{Bueler.Brown.2009, Balay.etal.2015} have
allowed for high-resolution numerical modelling of glaciers and ice sheets on
mountainous terrain
over millennial time scales \citep[e.g.,][]{Golledge.etal.2012}. Second, the
complex topography of the North American Cordillera also induces strong
geographic variations in temperature and precipitation, thus requiring the use
of high-resolution climate forcing fields as an input to an ice sheet model
\citep{Seguinot.etal.2014}. However, evolving climate conditions over the last
glacial cycle are subject to considerable uncertainty and still lie beyond the
computational reach of atmosphere circulation models.

@@

Ice rheology depends on temperature and water content through an enthalpy
formulation \citep{Aschwanden.etal.2012}. Surface air temperature derived from
the climate forcing (Sect.~\ref{sec:atm}) provides the upper boundary
condition to the ice enthalpy model. Temperature is computed subglacially to
a~depth of 3\,km,
where it is conditioned by a lower boundary geothermal heat flux of
70\,\unit{mW\,m^{-2}}. Although this uniform value does not account for the
high spatial geothermal variability in the region
\citep{Blackwell.Richards.2004}, it is, on average, representative of available
heat flow measurements. In the low-resolution simulations, the vertical grid
consists of 31~temperature layers in the bedrock and up to 51~enthalpy layers
in the ice sheet, corresponding to a vertical resolution of 100\,m. The
high-resolution simulations 61~bedrock layers and up to 101~ice layers with a
vertical resolution of 50\,m.

A~pseudo-plastic sliding law,
\begin{equation}
    \label{eqn:pseudoplastic}
    \vec{\tau}_b = -\tau_c \frac{\vec{v}_b}{{v_{th}}^q\,|\vec{v}_b|^{1-q}} \,,
\end{equation}
relates the bed-parallel shear stresses, $\vec{\tau}_b$, to the sliding
velocity, $\vec{v}_b$ (Table~\ref{tab:params}).
The yield stress, $\tau_c$,
is modelled using the Mohr--Coulomb criterion,
\begin{equation}
   \tau_c = c_0 + N\,\tan{\phi} \,,
\end{equation}
where cohesion, $c_0$, is assumed to be zero. The friction angle, $\phi$,
varies from 15 to 45\degree as a piecewise-linear function of modern bed
elevation, with the lowest value occuring below modern sea level (0\,m above
sea level, m~a.s.l.) and the highest value occuring above the generalised
elevation of the highest shorelines
\citep[200\,m~a.s.l.,][Fig.~5]{Clague.1981}, thus accounting for
a~weakening of till associated with the presence of marine sediments
(cf. \citealp{Martin.etal.2011}; \citealp[supplement]{Aschwanden.etal.2013};
\citealp{PISM-authors.2015}). Effective pressure, $N$,
is related to the ice overburden stress, $\rho gh$, and the modelled amount of
subglacial water, using a formula derived from laboratory experiments with till
extracted from the base of Ice Stream B in West Antarctica
\citep{Tulaczyk.etal.2000, Bueler.Pelt.2014},
\begin{equation}
    N = \delta \rho gh \, 10^{(e_0/C_c) (1 - (W/W_{max}))} \,,
\end{equation}
where $\delta$ is chosen as 0.02, $e_0$ is a measured reference void ratio and
$C_c$ is a measured compressibility coefficient (Table~\ref{tab:params}). The
amount of water at the base, $W$, varies from zero to $W_{max}=2$\,m, a
threshold above which additional melt water is assumed to drain off
instantaneously. Finally, the bedrock topography responds to ice load
following a bedrock deformation model that includes local isostasy,
elastic lithosphere flexure and viscous astenosphere deformation in an infinite
half-space \citep{Lingle.Clark.1985,Bueler.etal.2007}.
A relatively low viscosity value of $\nu_m = 1\times10^{19}$\,\unit{Pa\,s} is
used for the astenosphere (Table~\ref{tab:params}) in accordance with the
results from regional glacial
isostatic adjustment modelling at the northern Cascadia subduction zone
\citep{James.etal.2009}.

Ice shelf calving is computed using a double criterion. First, a
physically-realistic calving flux is computed based on eigenvalues of the
horizontal strain rate tensor \citep{Winkelmann.etal.2011,
Levermann.etal.2012}. This allows floating ice to advance in confined
embayments, but prevents the formation of extensive ice shelves in the open ocean.
Second, floating ice thinner than 50\,m is systematically calved off. A subgrid
scheme by \citet{Albrecht.etal.2011} allows for a continuous migration of the
calving front. This formulation of calving has been applied to the Antarctic ice
sheet and has shown to produce a realistic calving front positon for many of
the present-day ice shelves \citep{Martin.etal.2011}.

@@

\begin{figure}
  \includegraphics{atm}
  \caption{Monthly mean near-surface air temperature, precipitation, and
           standard deviation of daily mean temperature (PDD SD) for January
           and July from the North American Regional Reanalysis
           \citep[NARR;][]{Mesinger.etal.2006}, used to force the surface mass
           balance (PDD) component of the ice sheet model. Note the
           strong contrasts in seasonality, timing of the precipitation peak,
           and temperature variability over the model domain, notably between
           coastal and inland regions.}
  \label{fig:atm}
\end{figure}

@@

The PDD computation accounts for stochastic temperature variations by assuming
a normal temperature distribution of standard deviation $\sigma$ aroung the
expected value $T_m$. It is expressed by an error-function formulation
\citep{Calov.Greve.2005},
\begin{equation}
    \label{eqn:calovgreve}
    \mathrm{PDD} = \int_{t_1}^{t_2} \mathrm{d}t
        \left[\frac{\sigma}{\sqrt{2\pi}}
                \exp\left({-\frac{T_{m}^2}{2\sigma^2}}\right)
              + \frac{T_{m}}{2} \, \mathrm{erfc}
                \left(-\frac{T_{m}}{\sqrt{2}\sigma}\right)\right] \,,
\end{equation}
which is numerically approximated using week-long sub-intervals. In order to
account for the effects of spatial and seasonal variations of temperature
variability \citep{Seguinot.2013}, $\sigma$ is computed from NARR daily
temperature values from 1979 to 2000 \citep{Mesinger.etal.2006},
including variability associated with the seasonal cycle (Fig.~\ref{fig:atm}).
Degree-day factors for snow and ice melt are derived from
mass-balance measurements on contemporary glaciers from the Coast Mountains and
Rocky Mountains in British Columbia
\citep[Table~\ref{tab:params};][]{Shea.etal.2009}.

@@

\begin{table*}[t]
  \caption{Palaeo-temperature proxy records and scaling parameters yielding
           temperature offset time-series used to force the ice sheet model
           through the last glacial cycle (Fig.~\ref{fig:timeseries}). $f$
           corresponds to the scaling factor adopted to yield last glacial
           maximum ice limits in the vicinity of mapped end moraines, and
           $T_{[32, 22]}$ refers to the resulting mean temperature anomaly
           during the period -32 to~-22~ka after scaling.}
  \label{tab:records}
  {\begin{tabular}{l|ccc|ccc|l}
    \tophline
    Record & Latitude & Longitude & Elev. & Proxy & $f$ & $T_{[32, 22]}$
           & Reference\\
    & & & (m~a.s.l.) & & & (K) & \\
    \middlehline
    GRIP     &  72{\degree} 35' N  % 72.58 (decimal)
             &  37{\degree} 38' W  % 37.64 (decimal)
             & 3238
             & \chem{\delta^{18}O}
             & 0.37 & -5.8  % -16.4126 (before scaling)
             & \citet{Dansgaard.etal.1993} \\

    NGRIP    &  75{\degree} 06' N  % 75.10
             &  42{\degree} 19' W  % 42.32
             & 2917
             & \chem{\delta^{18}O}
             & 0.24 & -6.5  % -26.7098
             & \citet{Andersen.etal.2004} \\

    EPICA    &  75{\degree} 06' S  % 75.1
             & 123{\degree} 21' E  % 123.35
             & 3233
             & \chem{\delta^{18}O}
             & 0.64 & -5.9  % -9.2055
             & \citet{Jouzel.etal.2007} \\

    Vostok   &  78{\degree} 28' S  % 78.8
             & 106{\degree} 50' E  % 106.8
             & 3488
             & \chem{\delta^{18}O}
             & 0.74 & -5.9  % -7.9550
             & \citet{Petit.etal.1999} \\

    ODP~1012 &  32{\degree} 17' N
             & 118{\degree} 23' W
             & -1772
             & \chem{U^{K'}_{37}}
             & 1.61 & -6.1  % -3.7889
             & \citet{Herbert.etal.2001} \\

    ODP~1020 &  41{\degree} 00' N
             & 126{\degree} 26' W
             & -3038
             & \chem{U^{K'}_{37}}
             & 1.20 & -6.0 % -5.0000
             & \citet{Herbert.etal.2001} \\
    \bottomhline
  \end{tabular}}
  \belowtable{}
\end{table*}

@@

While simulations driven by the GRIP and the two Antarctic palaeo-temperature
records attain a last ice volume maximum between 19.1 and 16.8\,ka, those
informed by the NGRIP and the two oceanic palaeo-temperature records attain their
maximum ice volumes thousands of years earlier. Moreover, the ODP~1012 run
yields a rapid deglaciation of the modelled area prior to 17\,ka. The ODP~1020
simulation predicts an early maximum in ice volume at 23.2\,ka, followed by
slower deglaciation than modelled using the other palaeo-temperature records.
Finally, whereas model runs forced by the Antarctic palaeo-temperature records
result in a rapid and uninterrupted deglaciation after the LGM, the
simulation driven by the GRIP palaeo-temperature record also results in a rapid
deglaciation but in three steps, separated by two periods of ice sheet regrowth
(Fig.~\ref{fig:timeseries}).

@@

In the MIS~3 ice volume minimum reconstructions, a central ice cap persists
over the Skeena
Mountains (Fig.~\ref{fig:snapshots}, middle panels). Although this ice cap is
present in all simulations, its dimensions depend sensitively on the choice of
the applied palaeo-temperature record. Modelled ice volume minima spread over a
wide range between 1.69 and 2.88\,m~s.l.e. (Table~\ref{tab:extrema}).

Modelled ice sheet geometries during the LGM (MIS~2; Fig.~\ref{fig:snapshots},
lower panels) invariably include a ca. 1500\,km-long central divide above
3000\,m~a.s.l. located along the spine of the Rocky Mountains. Although the
similarity of modelled ice extents is a direct result of the choice of scaling
factors applied to different palaeo-temperature proxy records
(Table~\ref{tab:records}), it is interesting to note that modelled maximum ice
volumes also fall within a tight range of 8.40 to 8.91\,m~s.l.e.
(Table~\ref{tab:extrema}).

@@

Independently of the palaeo-temperature records
used to force the ice sheet model, our simulations consistently produce two
glacial maxima during the last glacial cycle. The first maximum configuration
is obtained during MIS~4 (61.9--56.5\,ka) and the second during MIS~2
(23.2--16.8\,ka; Figs.~\ref{fig:timeseries}, \ref{fig:snapshots};
Table~\ref{tab:extrema}). These events broadly correspond in timing to the
Gladstone
(MIS~4) and McConnell (MIS~2) glaciations documented by geological evidence for
the northern sector of the Cordilleran ice sheet
    \citep{Duk-Rodkin.etal.1996, Ward.etal.2007,
           Stroeven.etal.2010, Stroeven.etal.2014},
and to the Fraser Glaciation (MIS~2) documented for its southern sector
    \citep{Porter.Swanson.1998, Margold.etal.2014}.
There is stratigraphical evidence for an MIS~4 glaciation in British Columbia
\citep{Clague.Ward.2011} and in the Puget Lowland \citep{Troost.2014}, but
their extent and timing are still highly conjectural
    \citep[perhaps MIS~4 or early MIS~3; e.g.,][]{Cosma.etal.2008}.

@@

During MIS~2, the modelled total ice volume peaks at 8.01\,m~s.l.e. (19.1\,ka)
in the GRIP simulation and at 8.77\,m~s.l.e. (17.2\,ka) in the EPICA
simulation.
